\chapter{Algebra 2}
In this chapter we are going to learn how to solve a small group of linear equations.  The equations we will be looking at will consist of an unknown value and two sides.  With an equation the two sides must have the same value.

The equations that we look at will only deal with addition, subtraction, multiplication and division.

The types of equations that we will be looking at are:

\begin{multicols}{5}
$ax=b$

$\displaystyle \frac{x}{a}=b$

$x \pm a = b$

$ax \pm b = c$

$\displaystyle \frac{x}{a} \pm b = c$
\end{multicols}

\section{Opposites}
If we do something and then we do the exact opposite it is as if we have done nothing at all.  So if we were to walk 10km north and then walk 10km south we would end up where we started.

\begin{itemize}
	\item The opposite of adding is subtracting
	\item The opposite of subtracting is adding
	\item The opposite of multiplying is dividing
	\item The opposite of dividing is multiplying
\end{itemize}

\begin{exmp}
	Write down the opposite of the following:
	\begin{enumerate}[a)]
		\item Give \$10
		\item Walk 5km East
		\item +10
		\item $\div 6$
		\item -4
		\item $\times 7.3$
		\item $+5$ then $\times 4$
	\end{enumerate}
\bigskip

\noindent Answers

\begin{enumerate}[a)]
	\item Take \$10
	\item Walk 5km West
	\item -10
	\item $\times 6$
	\item +4
	\item $\div 7.3$
	\item $\div 4$ then $-5$
\end{enumerate}
\end{exmp}
\subsection{Exercise}
Write down the opposite of the following opperations
\begin{multicols}{3}
\begin{enumerate}
	\item Open door
	\item +6
	\item $\times 9$
	\item $\div 10$
	\item Go up 10m
	\item Go forward 7
	\item -3
	\item $+5$ then $\times 4$
	\item $\times 4$ then $+5$
\end{enumerate}
\end{multicols}
\section{One stage equations}
We will define a one stage a equation, an equation where we have either added or subtracted an number to our unknown value or multiplied or divided our unknown value by a number.  The four equations below will show you all four kinds of questions we are interested in:
\begin{multicols}{4}
\begin{itemize}
	\item $x+4=10$
	\item $x-7=3$
	\item $3x=12$
	\item $\displaystyle\frac{x}{7}=4$
\end{itemize}
\end{multicols}

When we are solving any kind of equation we must realise that if we are to keep both sides of the equation equal then if we change the size of one side, then we must change the size of the other side by an equal amount.  To solve our one stage problems we will follow 3 steps:

\begin{description}
	\item [Step 1] See what has been done to the unknown
	\item [Step 2] Do the opposite to both sides of the equation
	\item [Step 3] Simplify
\end{description}
\begin{exmp}

Solve:

$x + 5 = 12$

We need to cancel the $+5$ so we will subtract 5 from both sides.

So

$x + 5 -5 = 12 -5 $

$x = 7$

\noindent Solve:

$x - 4 = 9$

We need to cancel the $-4$ so we will add 4 to both sides.

$x = 13$

It is okay to miss that middle line provided we make it clear what we are doing.

\noindent Solve:

$5x = 30$

We need to cancel $\times 5$ so we will divide both sides by 5

$x = 6$

\noindent Solve:

\bigskip

$\displaystyle \frac{x}{3} = 9$

\bigskip
We need to cancel $\div 3$ so we will multiply by 3 on both sides.

$x = 27$
\end{exmp}
\subsection{Exercise}
Solve the following equations.  Make sure that you indicate what you are doing to solve the equation.
\begin{multicols}{3}
\begin{enumerate}
	\item $3x = 12$
	\item $x - 5 = 21$
	\item $5x = 35$
	\item $\displaystyle \frac{x}{2} = 6$
	\item $x + 7 = 9$
	\item $\displaystyle \frac{x}{4} = 5$
	\item $7x = 21$
	\item $x + 4.5 = 7.3$
	\item $2x = 9$
\end{enumerate}
\end{multicols}
\section{Two stage problems}
We will consider two stage problems where the unknown has been first multiplied or divided by a value and then a value is added or subtracted.  The form of the equations will be:

\begin{multicols}{2}
	$ax \pm b = c$
	
	$\displaystyle \frac{x}{a} \pm b = c$
\end{multicols}

\noindent To solve these problems we must first find the value of the term with the unknown.  we do this by cancelling out the value we have added or subtracted, we will then end up with a one stage problem.

\begin{exmp}

Solve $3x + 7 = 19$

First we will subtract 7 from both sides.

$3x = 12$

Now we divide both sides by 3.

$x = 4$
\end{exmp}

\begin{exmp}

\bigskip

Solve $\displaystyle \frac{x}{6} - 5 = 2$

\bigskip

First we should add 5 to both sides.

\bigskip

$\displaystyle \frac{x}{6} = 7$

\bigskip

Now we multiply both sides by 6.

$x = 42$
\end{exmp}

\subsection{Exercise}
In each of the equations below find the value of the unknown:
\begin{multicols}{2}
\begin{enumerate}
	\item $3x + 7 = 22$
	\item $2t - 5 = 13$
	\item $5p + 6 = 16$
	\item $\displaystyle \frac{f}{3} - 2 = 7$
	\item $\displaystyle \frac{x}{5} + 4 = 5$
	\item $2a + 3.5 = 4.7$
	\item $10b - 0.6 = 5.2$
	\item $\displaystyle \frac{q}{4} - 2 = 3.5$
	\item $\displaystyle \frac{y}{2} + 7.5 = 19$
	\item $5x-6 = 34$
	\item $2x + 3 = 10$
	\item $\displaystyle \frac{d}{8} + 5 = 7$
	\item $9g - 4 = 7$
	\item $\displaystyle \frac{p}{3} - 2 = 6$
	\item $\displaystyle \frac{l}{4} + 5 = 17$
\end{enumerate}
\end{multicols}